---
title: "The Problem"
author: "John Sterrett"
date: "2023-09-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
print(getwd())
source("../R/Cauda.R")
source("../R/AllClasses.R")

library(ggplot2)
library(magrittr)
```

#

```{r simulate_func, include=FALSE}

make_simple_confounded_dataset <- function(nsamples=1000,
                                           c_on_x=0.5,
                                           x_on_y=0.5,
                                           c_on_y=0.4){
  set.seed(42)
  # confounder
  tax.c <- rnbinom(n=nsamples, size=200, prob=0.3)
  # taxon X (independent var)
  tax.x <- c_on_x*tax.c + (1-c_on_x)*rnbinom(n=nsamples, size=200, prob=0.3)
  # matrix of taxa
  count.table <- matrix(data=c(tax.x, tax.c), nrow=nsamples)
  colnames(count.table) <- c("tax.x", "tax.c")

  # Y (dependent var)
  y <- x_on_y*tax.x +
       c_on_y*tax.c +
       (1-x_on_y-c_on_y)*rnorm(nsamples, mean=mean(tax.c), sd=sd(tax.c))

  # edge matrix to show that tax.c has a causal effect on tax.x
  edge.dictionary <- matrix(c(0,         0,
                              c_on_x!=0, 0),
                            nrow=2,
                            byrow=T)
  rownames(edge.dictionary) <- c("tax.x", "tax.c")
  colnames(edge.dictionary) <- c("tax.x", "tax.c")

  new("CauDAbox",
      metadata=matrix(data=y, nrow=nsamples),
      tableX=count.table,
      dictionary=edge.dictionary
      )
}

```

# Let's simulate some confounded data 
In this case, there is no causal effect of taxon x on the outcome, 
but it appears that there is due to confounding due to ecological interactions.

- Our taxon "of interest" has no effect on the outcome
- The confounding taxon has a negative effect on the taxon "of interest" (e.g., through direct predation)
- The confounding taxon has a positive effect on the outcome

```{r simulate}
# set some params
x.on.y <- 0  # Our taxon "of interest" has no effect on the outcome y
c.on.x <- -0.9 # The confounding taxon has a negative effect on the taxon "of interest"
c.on.y <- 0.5  # The confounding taxon has a positive effect on the outcome y

# make a confounded dataset for testing
confounded.data <- make_simple_confounded_dataset(1000,
                                       c_on_x=c.on.x,
                                       x_on_y=x.on.y,
                                       c_on_y=c.on.y)

colnames(confounded.data@metadata) <- c("depression")
#confounded.data@metadata <- scale(confounded.data@metadata, scale=10)
```

# Plotting confounded data {.tabset}
## Depression ~ Taxon X
```{r plot_confounded}
data <- cbind(confounded.data@metadata,
             confounded.data@tableX)
data <- as.data.frame(data)
ggplot(data,
       mapping=aes(y=depression, x=tax.x)) +
  geom_point() + 
  geom_smooth(method="lm",) +
  labs(x="Taxon X", y="Depression score") +
  theme_bw()
```

## Depression ~ Taxon C
```{r plot_confounder}
data <- cbind(confounded.data@metadata,
              confounded.data@tableX)
data <- as.data.frame(data)
ggplot(data,
       mapping=aes(y=depression, x=tax.c)) +
  geom_point() + 
  geom_smooth(method="lm",) +
  labs(x="Taxon C", y="Depression score") +
  theme_bw()
```

# Modeling the confounded data 

## Depression ~ Taxon X only
We can see below that because of the direct predation of taxon<sub>x</sub> by taxon<sub>c</sub>,
it appears that taxon<sub>x</sub> is causing lower depression scores. 
```{r}
broom::tidy(lm(depression ~ tax.x, data=data)) %>%
  knitr::kable()
```

## Depression ~ Taxon C only
In this case, taxon<sub>c</sub> has the same coefficient that we specified earlier, unlike taxon<sub>x</sub>.
```{r}
broom::tidy(lm(depression ~ tax.c, data=data)) %>%
  knitr::kable()
```

# The Solution?
## Regressing Depression ~ Taxon X + Taxon C
If we include the confounder in a regression with the taxon of interest, we recapture the actual effects of taxon<sub>x</sub> and taxon<sub>c</sub>.
```{r}
broom::tidy(lm(depression ~ tax.x + tax.c, data=data)) %>% 
  knitr::kable()
```

## Visualizing the non-confounded effects of Taxon X
If we regress out the effects of taxon<sub>c</sub> on taxon<sub>x</sub>, 
we also see the correct effect of Taxon X on depression score, and we can visualize that!

```{r}
resids.tax.x <- residuals(lm(tax.x ~ tax.c, data=data))

ggplot(mapping=aes(y=data$depression, x=resids.tax.x)) +
  geom_point() + 
  geom_smooth(method="lm",) +
  labs(x="Adjusted Taxon X", y="Depression score") +
  theme_bw()

broom::tidy(lm(depression ~ resids.tax.x, data=data)) %>% 
  knitr::kable()

```

# Final Thoughts
The issue is not that there are confounders. We know how to control for confounders!
In fact, many differential abundance tests (such as ANCOM-BC) allow users to specify
their own model, which can include confounders. However, in differential abundance 
packages, it is assumed that the confounders provided are relevant for every single 
relationship (y ~ taxon<sub>x</sub> + covariate) being tested. This doesn't allow
any room for taxon-specific ecological interactions or taxon-specific confounders.

Of course, this could be done manually, but no single researcher should have to 
manually specify every confounding ecological interaction within their 
microbiome dataset (and I would posit that nobody can know all this info).





